---
title: "ass2msb104grp2"
format: html
editor: visual
---

```{r}
#| label: setup
#| include: FALSE
library (tidyverse)
library (PxWebApiData)
library(readxl)
library(purrr)
library(dineq)
library(psych)
library(flextable)
```

## Our chosen variabels 
In part B we have chosen the variables: Labour force, Unemployment and Population density. We will look in to each of these variable, and see how they are a potential a determaint factor for the regional inequality. First lets look at what Regional inequality, and it is the variable that describes an uneven distribution of resources, opportunities and well-being across geographic areas.
Labour force is the variable which includes all persons how are both employed and unemployed. It represent the total supply of labour that's available in the regions. Labour force can factor in the different between regions avabile workforce which then affect the regions inequality.
The Unemployment rate variable can be seen as a sub sector from labour force, as these are persons without work, available for work and are seeking work. The variable is shown as a percentage of the labour force. With this variable we may look into if the The unemployment rate has an effect on the income of those in the region, which affects GDP, which will then have a effect on the regional inequality.
The last is the population density in each region. This variable shows the relationship between the annual average population and the land area of the respective region. With this variable we can see the different in population by area of each region, which may have a affect on regional inequality.


```{r}
#Removing the metadata from the top and bottom by defining the range.
raw_Labourforce <- read_excel("./Data/Labour force by NUTS2.xlsx", sheet = "Sheet 1", range = "A10:C66", col_types = "text")
#Dropping the first row to align time with years
   raw_Labourforce <- raw_Labourforce[-1,]
   names(raw_Labourforce)[1] <- "Geo_Codes"
   names(raw_Labourforce)[2] <- "Geo_Labels"
as_flextable(raw_Labourforce)
```


```{r}
# Making raw_Labourforce into long format.
tidy_Labourforce <- pivot_longer(data = raw_Labourforce,
             cols = c(-Geo_Codes, -Geo_Labels),
             names_to = "Time",
             values_to = "Thousand persons")
```


```{r}
#Removing the metadata from the top and bottom by defining the range.
raw_Unemployment <- read_excel("./Data/Unemployment rate NUTS2.xlsx", sheet = "Sheet 1", range = "A11:C68", col_types = "text")
#Dropping the first row to align time with years
   raw_Unemployment <- raw_Unemployment[-1,]
   names(raw_Unemployment)[1] <- "Geo_Codes"
   names(raw_Unemployment)[2] <- "Geo_Labels"
as_flextable(raw_Unemployment)
```

```{r}
# Making raw_Unemployment into long format.
tidy_Unemployment <- pivot_longer(data = raw_Unemployment,
             cols = c(-Geo_Codes, -Geo_Labels),
             names_to = "Time",
             values_to = "Percentage")
```


```{r}
#Removing the metadata from the top and bottom by defining the range.
raw_Population.density <- read_excel("./Data/Population density NUTS2.xlsx", sheet = "Sheet 1", range = "A8:C60", col_types = "text")
#Dropping the first row to align time with years
   raw_Population.density <- raw_Population.density[-1,]
   names(raw_Population.density)[1] <- "Geo_Codes"
   names(raw_Population.density)[2] <- "Geo_Labels"
as_flextable(raw_Population.density)
```

```{r}
# Making raw_Population.density into long format.
tidy_Populatiopn.density <- pivot_longer(data = raw_Population.density,
             cols = c(-Geo_Codes, -Geo_Labels),
             names_to = "Time",
             values_to = "Persons per square kilometre")
```



```{r}
#| label: tbl-tidyjoined
#| fig-cap: "Slått sammen og ryddet datasett med NUTS2 som kolonne nummer 2"

# Slår sammen tre datasett basert på Geo_Labels og Time
tidyjoined2 <- list(tidy_Populatiopn.density, tidy_Labourforce, tidy_Unemployment) %>%
  reduce(left_join, by = join_by(Geo_Labels, Time)) %>%
  select(-matches("\\.x$"), -matches("\\.y$"))  # fjerner duplikate kolonner

# Rydder og setter riktig kolonnerekkefølge
tidyjoined2 <- tidyjoined2 %>%
  mutate(
    `Persons per square kilometre` = as.numeric(`Persons per square kilometre`),
    Percentage = as.numeric(Percentage),
    `Thousand persons` = as.numeric(`Thousand persons`),
    NUTS2 = str_sub(Geo_Codes, start = 1L, end = 4L)
  ) %>%
  select(NUTS2, Geo_Labels, Time,
         `Persons per square kilometre`,
         `Thousand persons`,
         Percentage)  # NUTS2 som kolonne nr 2

# Viser tabellen pent
as_flextable(tidyjoined2)

```






