% Options for packages loaded elsewhere
% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  british,
  a4paper,
]{article}
\usepackage{xcolor}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{5}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother


\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother


% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
\ifLuaTeX
  \usepackage[english]{selnolig} % disable illegal ligatures
\fi


\setlength{\emergencystretch}{3em} % prevent overfull lines

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}



 


\usepackage{fontspec}
\usepackage{multirow}
\usepackage{multicol}
\usepackage{colortbl}
\usepackage{hhline}
\newlength\Oldarrayrulewidth
\newlength\Oldtabcolsep
\usepackage{longtable}
\usepackage{array}
\usepackage{hyperref}
\usepackage{float}
\usepackage{wrapfig}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Assignment 4:},
  pdfauthor={Kristoffer Tufta and Harald Vika},
  pdflang={en-GB},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}


\title{Assignment 4:}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{Regional GDP Inequality in 4 Selected European Economies -
Synthesised Results.}
\author{Kristoffer Tufta and Harald Vika}
\date{Wednesday 3 Dec, 2025}
\begin{document}
\maketitle


\section{Assignment 4 Consolidation of Key
findings}\label{assignment-4-consolidation-of-key-findings}

\subsection{Abstract}\label{abstract}

\subsection{Introduction}\label{introduction}

\subsubsection{Background}\label{background}

\subsubsection{Objectives}\label{objectives}

\subsubsection{Significance}\label{significance}

\subsection{Literature Review -
Kristoffer}\label{literature-review---kristoffer}

\subsubsection{Previous Work}\label{previous-work}

A central reference point for our work is Lessmann \& Seidel (2017), who
shows that the link between economic development and regional inequality
is not linear. Their results suggest something closer to an inverted-U
(and in some cases N-shaped) pattern, where inequality tends to rise in
the middle stages of development before falling again in high-income
settings. This gives a natural expectation that fast-growing regions may
show lower inequality if they are already relatively advanced, while
poorer or developing regions might move in the opposite direction. Their
work is also relevant because they rely on both traditional data and
satellite-based luminosity data, which shows that inequality can be
meaningfully captured even when official data is patchy --- a recurring
issue even in Eurostat datasets, as we experienced directly.

Eurostat's own methodological documentation is also part of the relevant
background. Their notes on missing values help explain why regional GDP
and population numbers are incomplete in certain years. These gaps arise
from reclassification of regions, confidentiality rules, or delayed
reporting. Understanding why these ``quirks'' happen is important
because it directly affects our results.

For our segmentation work, we also build on Eurostat's definitions of
population-density classes used for identifying rural areas, urban
clusters, and high-density urban centres. While Eurostat's approach
operates at the 1 km² continuous grid-cell level, their breakpoints (300
and 1500 people per square kilometer) give a practical structure for
grouping NUTS2 regions by density. This matters because urban density is
often correlated with both productivity and access to public-services,
and therefore a likely influence when it comes to regional GDP
inequality.

\subsubsection{Research Gap}\label{research-gap}

While there is existing literature on national GDP inequality trends and
on broader development patterns, there is less focus on how short-term
growth shocks play out inside individual NUTS2 regions, and even less on
whether the effect of growth differs across structural categories like
urban density, unemployment rates, or workforce size. Our assignments
aim to examine exactly that: to see whether any of these structural
categories influence GDP inequality in NUTS2-regions, and if so, to what
degree.

\subsection{Data and Methodology}\label{data-and-methodology}

\subsubsection{Data Sources - Harald}\label{data-sources---harald}

Primary data refers to information collected firsthand by the researcher
directly from original source. It is raw and unprocessed, typically
gathered through methods such as surveys, interviews and observations.
The data provided by Eurostat can be considered primary data because it
is originally collected through official surveys and processed following
established methodological standards for gathering information on
specific topic within defined time periods.

In the first assignment, we used Eurostat to extract datasets on gross
domestic product (GDP) at current market prices by NUTS 3 regions. From
these datasets, we collected data for Germany, Ireland, Croatia annd
Switzerland covering the years 2000 to 2023.

In the second assignment, we retrieved data from three additional
datasets that could potentially influence regional inequality in 2017.
These datasets included labour force by NUTS 2 region, unemployment rate
by Nuts 2 region, and population density by NUTS 2 region. Each country
collects its own data and reports it to Eurostat in accordance with ESA
2010, a global framework for national accounting. This ensures that the
resulting statistics are consistent, reliable, and comparable across
countries.

However, the data also comes with certain limitations. For example,
Switzerland is not en EU member but an EFTA country (European Free Trade
Association), which promotes free trade and economic cooperation.
Because Switzerland did not have a formal datasharing agreement with
Eurostate before 2008, some data for earlier years or specific topics
may be missing (\emph{Information on Data - National Accounts -
Eurostat}, n.d.).

Another limitation relates to confidentiality. Ireland, for instance,
lacks GDP data for 2015-2017 in the Mid-West and South-West regions due
to confidentiality restrictions imposed by Ireland's Central Statistics
Office. These restrictions are linked to changes in the national
accounts in 2015 that significantly affected the measurement of
Ireland's productive capacity (\emph{County Incomes and Regional GDP
2015 - CSO - Central Statistics Office}, 2018).

Some counters also choose to delay the publication of certain statistics
to ensure accuracy. As a result, data for the most recent years of 2022
and 2023, may still be incomplete due to its unavailability.

A common limitation across several datasets concerns changes in regional
boundaries. Regions may merge to form larger areas or split into smaller
units, affecting how data is collected, aggregated and updated. This
issue is particularly relevant in Germany, where regional reforms occur
more frequntly due to regional policy. Coratia and Ireland face similar
challenges, as changes to their NUTS regions have affected population
data availability at the NUTS 3 level.

Finally, there were limitations related to the time coverage of the
datasets. Althought the assignments aimed to examine data from 2000 to
2023, the population density dataset for NUTS 2 regions only covers
2012-2023, and the unemployment rate dataset covers 2013-2023.
Consequently, when including these variables, we were restricted to
using data from 2013 to 2023 to maintain consistency across datasets.

\subsubsection{Methodological approach -
Kristoffer}\label{methodological-approach---kristoffer}

\paragraph{Cross-Sectional Estimation}\label{cross-sectional-estimation}

We first combined the Eurostat datasets covering GDP (in million euros)
and population, , then calculated GDP per capita and Gini coefficients
using a formula weighted by population, like Lessmann \& Seidel (2017).
We then visualized the calculated Gini coefficients for the different
regions using different plots from ggplot2. In assignment 2, we made a
cross-sectional analysis of the year 2017 using a regression model to
test the effect of change in regional GDP per capita on regional
inequality. We specified the following linear regression model: \[
\text{Gini}_{i,2017} = \alpha + \beta \, \Delta \text{GDPpc}_{i,2016 \to 2017} + u_i
\] Where:

\begin{itemize}
\tightlist
\item
  \(\text{Gini}_{i,2017}\) --- Calculated Gini coefficient of region
  \emph{i} in 2017 (regional GDP per capita inequality)\\
\item
  \(\Delta \text{GDPpc}_{i,2016 \to 2017}\) --- percent change in GDP
  per capita from 2016 to 2017 in region \emph{i}\\
\item
  \(\alpha\) --- intercept term (baseline inequality when GDP-per-capita
  change is zero)\\
\item
  \(\beta\) --- slope coefficient showing how inequality changes with a
  one-percentage-point increase in GDP-per-capita growth\\
\item
  \(u_i\) --- error term capturing unobserved regional factors
\end{itemize}

We then experimented with other determinants of inequality by making a
new OLS model with a set of new variables: population density (persons
per square kilometer), unemployment rate and total workforce, also
fetched from Eurostat. We specified the MLR as follows: \[
\text{Gini}_{i,2017} = \alpha + \beta_1\,\text{Workforce}_i + \beta_2\,\text{PopDensity}_i + \beta_3\,\text{Unemployment}_i + u_i
\]

where \(\alpha\) is the intercept, \(\beta_1\), \(\beta_2\), and
\(\beta_3\) are the slope coefficients for each explanatory variable,
and \(u_i\) is the error term capturing unobserved factors affecting
inequality.

\paragraph{Segmentation}\label{segmentation}

We also segmented regions into categories to examine if the
development--inequality relationship differed across different
structural environments. Population density was split into ``Rural'',
``Medium Density'', and ``High Density'' using Eurostat-inspired
breakpoints (0--300, 300--1500, 1500+). For workforce size, we used
quantile-based grouping (Low, Medium, High). Unemployment rate was also
split into Low-High segments, but here we set the breakpoints manually
at 0\%-5\%, 5\%-10\% and 10+\%. These subsets were used to run separate
regressions for each group.

\paragraph{Panel Estimation}\label{panel-estimation}

Finally, we estimated fixed-effects models using the full panel,
applying plm() with region, country, year, and two-way effects. Before
estimation, we converted the dataset into a pdata.frame and cleaned
remaining NA/Inf values to avoid dropped regions. The panel models
allowed us to control for all time-invariant regional characteristics,
giving a ``less noisy'' picture of how regional GDP inequality changes
inside regions over time.

\paragraph{Handling of NA-values and
Heteroskedasticity}\label{handling-of-na-values-and-heteroskedasticity}

After joining the datasets onto eachother, including the new variables
used for segmenting, we ended up with a few obvious gaps in our data and
some NA/Inf rows. These were removed to keep the panel consistent, but a
side effect of this is that we only have panel data from 2013 onwards.

We also ran Breusch--Pagan tests to check for heteroskedasticity across
all regressions. Where alternative functional forms were required, we
introduced quadratic and logarithmic transformations of the key
variables and evaluated their residual behaviour the same way.

\subsection{Empirical Findings -
Harald}\label{empirical-findings---harald}

\subsubsection{Cross- sectional
Estimates}\label{cross--sectional-estimates}

In assignment 2, we began with a cross-sectional analysis examining the
relationship between regional inequality and short-term ecnonmic
development, measured as the change in GDP per capita from 2016 to 2017.
The dependent variable in our model was regional inequality, measured by
the Gini coefficent, and the independent variable was the change in GDP
per capita. This produced the following estimates:

\global\setlength{\Oldarrayrulewidth}{\arrayrulewidth}

\global\setlength{\Oldtabcolsep}{\tabcolsep}

\setlength{\tabcolsep}{2pt}

\renewcommand*{\arraystretch}{1.5}



\providecommand{\ascline}[3]{\noalign{\global\arrayrulewidth #1}\arrayrulecolor[HTML]{#2}\cline{#3}}

\begin{longtable*}[c]{|p{0.98in}|p{0.88in}|p{1.29in}|p{0.75in}|p{0.75in}|p{0.40in}}



\ascline{1.5pt}{666666}{1-6}

\multicolumn{1}{>{\raggedright}m{\dimexpr 0.98in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{Estimate}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.29in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{Standard\ Error}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{t\ value}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{Pr(>|t|)}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{}}}} \\

\ascline{1.5pt}{666666}{1-6}\endfirsthead 

\ascline{1.5pt}{666666}{1-6}

\multicolumn{1}{>{\raggedright}m{\dimexpr 0.98in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{Estimate}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.29in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{Standard\ Error}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{t\ value}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{Pr(>|t|)}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{}}}} \\

\ascline{1.5pt}{666666}{1-6}\endhead



\multicolumn{1}{>{\raggedright}m{\dimexpr 0.98in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{(Intercept)}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{0.134}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.29in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{0.016}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{8.514}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{0.0000}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{***}}}} \\





\multicolumn{1}{>{\raggedright}m{\dimexpr 0.98in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{chg\_gdpc}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{-0.002}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.29in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{0.003}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{-0.787}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{0.4350}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{\ \ \ }}}} \\

\ascline{1.5pt}{666666}{1-6}



\multicolumn{6}{>{\raggedleft}m{\dimexpr 5.05in+10\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{\textit{Signif.\ codes:\ 0\ <=\ '***'\ <\ 0.001\ <\ '**'\ <\ 0.01\ <\ '*'\ <\ 0.05}}}}} \\





\multicolumn{6}{>{\raggedright}m{\dimexpr 5.05in+10\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{}}}} \\





\multicolumn{6}{>{\raggedright}m{\dimexpr 5.05in+10\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{Residual\ standard\ error:\ 0.08068\ on\ 50\ degrees\ of\ freedom}}}} \\





\multicolumn{6}{>{\raggedright}m{\dimexpr 5.05in+10\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{Multiple\ R-squared:\ 0.01224,\ Adjusted\ R-squared:\ -0.007518}}}} \\





\multicolumn{6}{>{\raggedright}m{\dimexpr 5.05in+10\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{F-statistic:\ 0.6194\ on\ 50\ and\ 1\ DF,\ p-value:\ 0.4350}}}} \\





\end{longtable*}



\arrayrulecolor[HTML]{000000}

\global\setlength{\arrayrulewidth}{\Oldarrayrulewidth}

\global\setlength{\tabcolsep}{\Oldtabcolsep}

\renewcommand*{\arraystretch}{1}

The results shows an estimated intercept of (\(\hat{\alpha}=0.1336\)),
representing the average level of inequality in regions with zero
GDP-per-capita growth. The coefficient for GDP-per-capita growth is
negative (\(\hat{\beta}=-0.0024\)), which indicates a slight negative
relationship between economic growth and inequality. However, the
model's explanatory power is extremely low, with (\(R^2=0.012\)),
meaning that changes in GDP-per-capita account for only 1.2\% of the
variation in inequality. Thus, although the coefficient suggests a
negative relationship, the effect is statistically insignificant and not
substantively meaningful. We can conclude that the short-term
GDP-per-capita growth does has a small to none explanatory power for
regional inequality in this analyse.

We then went on to collect three more variables to determine whether
they serve as strong predictor of regional inequality then
GDP-per-capita growth. These variables were labour force, unemployment
rate, and population density. After cleaning and tiding the data, we
estimated a multiple regression model to analyse their cross-sectional
relationship with inequality: \[
\text{Gini}_{i,2017} = \alpha + \beta_1\,\text{Workforce}_i + \beta_2\,\text{PopDensity}_i + \beta_3\,\text{Unemployment}_i + u_i
\]

Where:

\begin{itemize}
\tightlist
\item
  \(\text{Gini}_{i,2017}\) --- Calculated Gini coefficient of region
  \emph{i} in 2017 (regional GDP per capita inequality)
\item
  \(\Delta \text{Workforce}_{i}\) --- Total labour force in region i,
  including both employed and unemployed individuals
\item
  \(\Delta \text{PopDensity}_{i}\) --- Population density of region i,
  indicating how densely populated the region is
\item
  \(\Delta \text{Unemployment}_{i}\) --- Unemployment rate in region i,
  representing presons withouth work who are available and is seeking
  work.
\item
  \(\alpha\) --- intercept term (baseline inequality when the variables
  change is zero)
\item
  \(\beta_1, \beta_2, \beta_3\) --- Coefficient showing how inequality
  changes with a one-percentage-point change in each variable
\item
  \(u_i\) --- error term capturing unobserved regional factors
\end{itemize}

Running the regression model gave us the following estimates:

\global\setlength{\Oldarrayrulewidth}{\arrayrulewidth}

\global\setlength{\Oldtabcolsep}{\tabcolsep}

\setlength{\tabcolsep}{2pt}

\renewcommand*{\arraystretch}{1.5}



\providecommand{\ascline}[3]{\noalign{\global\arrayrulewidth #1}\arrayrulecolor[HTML]{#2}\cline{#3}}

\begin{longtable*}[c]{|p{1.12in}|p{0.88in}|p{1.29in}|p{0.75in}|p{0.75in}|p{0.40in}}



\ascline{1.5pt}{666666}{1-6}

\multicolumn{1}{>{\raggedright}m{\dimexpr 1.12in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{Estimate}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.29in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{Standard\ Error}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{t\ value}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{Pr(>|t|)}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{}}}} \\

\ascline{1.5pt}{666666}{1-6}\endfirsthead 

\ascline{1.5pt}{666666}{1-6}

\multicolumn{1}{>{\raggedright}m{\dimexpr 1.12in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{Estimate}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.29in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{Standard\ Error}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{t\ value}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{Pr(>|t|)}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{}}}} \\

\ascline{1.5pt}{666666}{1-6}\endhead



\multicolumn{1}{>{\raggedright}m{\dimexpr 1.12in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{(Intercept)}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{0.150168}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.29in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{0.034}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{4.465}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{0.0001}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{***}}}} \\





\multicolumn{1}{>{\raggedright}m{\dimexpr 1.12in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{Workforce}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{0.000043}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.29in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{0.000}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{2.380}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{0.0216}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{\ \ *}}}} \\





\multicolumn{1}{>{\raggedright}m{\dimexpr 1.12in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{Pop\_km2}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{-0.000041}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.29in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{0.000}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{-2.763}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{0.0083}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{\ **}}}} \\





\multicolumn{1}{>{\raggedright}m{\dimexpr 1.12in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{Unemp\_prct}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{-0.011422}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.29in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{0.006}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{-1.919}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{0.0613}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{\ \ .}}}} \\

\ascline{1.5pt}{666666}{1-6}



\multicolumn{6}{>{\raggedleft}m{\dimexpr 5.19in+10\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{\textit{Signif.\ codes:\ 0\ <=\ '***'\ <\ 0.001\ <\ '**'\ <\ 0.01\ <\ '*'\ <\ 0.05}}}}} \\





\multicolumn{6}{>{\raggedright}m{\dimexpr 5.19in+10\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{}}}} \\





\multicolumn{6}{>{\raggedright}m{\dimexpr 5.19in+10\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{Residual\ standard\ error:\ 0.06949\ on\ 45\ degrees\ of\ freedom}}}} \\





\multicolumn{6}{>{\raggedright}m{\dimexpr 5.19in+10\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{Multiple\ R-squared:\ 0.2826,\ Adjusted\ R-squared:\ 0.2347}}}} \\





\multicolumn{6}{>{\raggedright}m{\dimexpr 5.19in+10\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\global\setmainfont{Arial}{F-statistic:\ 5.908\ on\ 45\ and\ 3\ DF,\ p-value:\ 0.0017}}}} \\





\end{longtable*}



\arrayrulecolor[HTML]{000000}

\global\setlength{\arrayrulewidth}{\Oldarrayrulewidth}

\global\setlength{\tabcolsep}{\Oldtabcolsep}

\renewcommand*{\arraystretch}{1}

This model gives us a R² of 0,283 and an adjusted R² of 0,235 meaning
that the these three variables together has a 28\% explanatory power of
the variation in regional inequality, which is higher then the change in
GDP-per-capita. The \emph{workforce} variable is statistically
significant at the 5\% level (p = 0.0216). It's coefficient,
\(\beta = 0.000043\), implies that regional inequality increases
slightly as the labour force grows by one thousand workers. This may
reflect that regions with larger or more active labour markets
experience greater occupational and wage diversity, leading to somewhat
higher inequality.

\emph{Population density} is statistically significant at the 1\% level
(p = 0.0083) and has a negative coefficient (\(\beta = -0.000041\)).
This suggests that more densely populated---typically more
urbanized---regions tend to have lower inequality, possibly due to
better access to education, more specialized employment opportunities,
and a broader labour market.

The \emph{unemployment} variable has a coefficient of
\(\beta = -0.0114\) and is marginally significant at the 10\% level (p =
0.0613). The negative sign indicates that regions with higher
unemployment rates appear to have lower inequality, although the weak
significance prevents us from drawing a firm conclusion.

\emph{Overall assessment}: Both labour force size and population density
are stronger predictors of regional inequality than short-term
GDP-per-capita growth. These structural variables appear to better
capture regional economic differences than year-to-year changes in
output.

\subsubsection{Alternative functional forms and panel
estimates}\label{alternative-functional-forms-and-panel-estimates}

In this section, we explored alternative functional forms of the
regression models using the variables from the second assignment
workforce, population density, and unemployment rate to examine their
effects on regional inequality, measured by the Gini coefficient.
Specifically, we considered linear, logarithmic, and quadratic
functional forms. While cubic specifications were also tested, they
produced more estimates without providing meaningful improvements in
explanatory power, so they were not pursued further.

We constructed separate regression tables for each functional form
linear, logarithmic, and quadraticand compared their performance in
terms of coefficient significance, explanatory power, and overall model
fit. Below, we summarize the key findings and interpret the effects of
each independent variable.

Workforce

Linear model:

\begin{itemize}
\item
  When all independent variables are zero, the regional inequality
  coefficient is estimated at \(\beta = 0.1773\).
\item
  Workforce has a coefficient of \(\beta = 0.000036\) (p = 0.0665),
  which is marginally significant at the 10\% level.
\item
  Change in GDP per capita has a coefficient of \(\beta = -0.002122\) (p
  = 0.4931), indicating it is not statistically significant.
\item
  Overall, the model's explanatory power is low: \(R^2 = 0.074\),
  adjusted \(R^2 = 0.035\), and the F-test (F = 1.877, p = 0.163)
  suggests the model is not statistically significant.
\end{itemize}

Logarithmic model:

\begin{itemize}
\tightlist
\item
  Transforming change in GDP per capita using a log transformation
  improves model performance.
\item
  When all independent variables are zero, the regional inequality
  coefficient is estimated at \(\beta = 0.0.2292\).
\item
  The coefficient for GDP per capita becomes \(\beta = -0.059611\) and
  is highly significant (p = 0.0008), showing a negative relationship
  with the Gini coefficient.
\item
  Workforce remains largely unchanged at \(\beta = 0.000033\) (p =
  0.0604, marginally significant).
\item
  Explanatory power improves: \(R^2 = 0.288\), adjusted
  \(R^2 = 0.0251\), and the F-test is highly significant (F = 8.078, p =
  0.0011).
\end{itemize}

Quadratic model:

\begin{itemize}
\tightlist
\item
  The quadratic specification produces less significant results than the
  logarithmic model.
\item
  When all independent variables are zero, the regional inequality
  coefficient is estimated at \(\beta = 0.1379\).
\item
  GDP per capita: linear term \(\beta = 0.004327\) (p = 0.4066),
  quadratic term \(\beta = -0.000468\) (p = 0.1302), both insignificant.
\item
  Workforce: \(\beta = 0.000029\) (p = 0.1480), effect is weak.
\item
  \(R^2 = 0.1194\), adjusted \(R^2 = 0.06199\), and the model is not
  statistically significant overall (p = 0.116).
\end{itemize}

Summary: The logarithmic model best captures the relationship between
workforce, GDP growth, and inequality, mainly due to the strong
significance of log-transformed GDP per capita.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\tightlist
\item
  Population Density
\end{enumerate}

Linear model:

\begin{itemize}
\tightlist
\item
  When change in GDP per capita and population density are zero, the
  regional inequality coefficient is estimated at \(\beta = 0.1506\).
\item
  Population density has a coefficient of \(\beta = -0.000037\) with a
  p-value of 0.0228, indicating statistical significance at the 5\%
  level.
\item
  Change in GDP per capita has a coefficient of \(\beta = 0.000054\) (p
  = 0.6166), which is not significant.
\item
  Model performance is limited: \(R^2 = 0.111\), adjusted
  \(R^2 = 0.072\), and the F-test is 7.36 with a p-value of 0.00668,
  suggesting marginal overall significance.
\end{itemize}

Logarithmic model:

\begin{itemize}
\tightlist
\item
  Applying a log transformation to change in GDP per capita improves
  model fit and significance.
\item
  When change in GDP per capita and population density are zero, the
  regional inequality coefficient is estimated at \(\beta = 0.229228\).
\item
  GDP per capita becomes \(\beta = -0.05793\), p = 0.0007, showing a
  strong negative relationship with inequality.
\item
  Population density remains similar: \(\beta = -0.000039\), p = 0.0047,
  still significant.
\item
  Overall explanatory power increases substantially: \(R^2 = 0.364\),
  adjusted \(R^2 = 0.331\), and the F-test p-value is 0.0001, indicating
  a highly significant model.
\end{itemize}

Quadratic model:

\begin{itemize}
\tightlist
\item
  Including a quadratic term for GDP per capita allows for a non-linear
  relationship with inequality.
\item
  When change in GDP per capita and population density are zero, the
  regional inequality coefficient is estimated at \(\beta = 0.137873\).
\item
  GDP per capita: linear term \(\beta = 0.0077\) (p = 0.1136, not
  significant), quadratic term \(\beta = -0.000682\) (p = 0.0198,
  significant), indicating a slight non-linear effect.
\item
  Population density: \(\beta = -0.000041\), p = 0.0091, remains
  significant.
\item
  Model fit is moderate: \(R^2 = 0.213\), adjusted \(R^2 = 0.161\), with
  overall model p = 0.0123, showing statistical significance.
\end{itemize}

Summary: The logarithmic model provides the best explanatory power for
population density, with both GDP growth and population density showing
significant impacts on regional inequality. The quadratic model captures
a slight non-linear effect but has lower overall explanatory power.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\tightlist
\item
  Unemployment Rate
\end{enumerate}

Linear model:

\begin{itemize}
\tightlist
\item
  The regional inequality coefficient is \(\beta = 0.1758\) when GDP per
  capita growth and unemployment are zero.
\item
  Unemployment rate: \(\beta = -0.001039\), p = 0.03, statistically
  significant, suggesting higher unemployment is associated with
  slightly lower inequality.
\item
  Change in GDP per capita: \(\beta = -0.0000248\), p = 0.936, not
  significant.
\item
  Model fit is weak: \(R^2 = 0.10\), adjusted \(R^2 = 0.062\), and the
  F-test p-value is 0.083, indicating borderline overall significance.
\end{itemize}

Logarithmic model:

\begin{itemize}
\tightlist
\item
  Transforming GDP per capita using a log improves model performance.
\item
  The regional inequality coefficient is \(\beta = 0.233016\) when GDP
  per capita growth and unemployment are zero.
\item
  GDP per capita: \(\beta = -0.051\), p = 0.005, now highly significant
  with a negative effect on inequality.
\item
  Unemployment rate: \(\beta = -0.00707\), p = 0.1096, not significant,
  though the effect remains negative.
\item
  Explanatory power increases: \(R^2 = 0.27\), adjusted
  \(R^2 = 0.2336\), F-test p = 0.0018, highly significant overall.
\end{itemize}

Quadratic model:

\begin{itemize}
\item
  Including a quadratic term for GDP per capita allows for a potential
  non-linear effect.
\item
  The regional inequality coefficient is \(\beta = 0.160671\) when GDP
  per capita growth and unemployment are zero.
\item
  GDP per capita: linear term \(\beta = 0.0062\) (p = 0.2114, not
  significant), quadratic term \(\beta = -0.000488\) (p = 0.0102,
  borderline significant).
\item
  Unemployment rate: \(\beta = -0.009216\), p = 0.0516, borderline
  significant.
\item
  Model fit is lower than the log model: \(R^2 = 0.1517\), adjusted
  \(R^2 = 0.09633\), with F-test p = 0.0539, borderline significant
  overall.
\end{itemize}

Summary: The logarithmic transformation of GDP per capita produces the
strongest model for unemployment, improving both explanatory power and
overall significance. Unemployment shows a weak but negative
relationship with inequality, while quadratic effects of GDP per capita
are borderline significant.

Overall Conclusion for the alternative functional forms:

Across all variables, logarithmic models generally provide the best
explanatory power and statistical significance, particularly through the
transformation of GDP per capita. Quadratic models capture some
non-linear effects but often at the cost of reduced overall
significance. Linear models are generally the weakest in explaining
regional inequality. The key takeaway is that GDP per capita growth
consistently shows a negative relationship with inequality when
appropriately transformed, while workforce, population density, and
unemployment rate exhibit more nuanced, context-dependent effects.

The panel regression results reveal a consistently positive association
between change in GDP per capita and the dependent variable across all
specifications. The magnitude of the effect is small (0.00017--0.00024),
but precision improves substantially when controlling for regional
heterogeneity using NUTS2 fixed effects, as reflected in markedly lower
standard errors and RMSE. Models with country only or year only fixed
effects explain very little variation (R² = 0.005--0.009), whereas NUTS2
and NUTS2 \& Year fixed effects substantially improve model fit (R² =
0.102--0.109) and are preferred according to AIC and BIC. These findings
highlight the importance of accounting for subnational regional
characteristics, which capture significant variation in the outcome
variable. Overall, the results suggest that regional-level factors are
key drivers of the dependent variable, while the positive effect of
change in GDP per capita remains robust across specifications.

\subsection{Discussion - Begynner kvar for oss, samskriver
etterpå}\label{discussion---begynner-kvar-for-oss-samskriver-etterpuxe5}

\begin{itemize}
\tightlist
\item
  Key insight:
\end{itemize}

The logarithmic models gives the best results for the alternative
functions.

\subsubsection{Key insights}\label{key-insights}

\subsubsection{Policy Implications}\label{policy-implications}

Across all the cross-sectional and panel models, the effect of GDP per
capita growth on inequality is small, unstable, and mostly
insignificant. For policymakers, this means that betting on regional
inequality improvements through ``more growth'' alone would be a
sub-optimal approach The differences between regions appear to be driven
far more by long-term structural characteristics than by year-to-year
economic performance.

The segmentation exercises support this interpretation, although with a
small asterisk attached. Regions with higher population density tend to
show slightly lower inequality, while rural regions display more
variation and, in some cases, higher inequality. This suggests that
sparsely populated areas face structural disadvantages---smaller labour
markets, fewer high-wage occupations, weaker public services---that
growth alone cannot compensate for. At the same time, the way
NUTS2-regions are defined (generally based on existing administrative
divisions), makes it so that we cannot take this at face value. This
division tends to separate city-regions into their own separate
NUTS2-regions, meaning that the most densely populated areas also by
definition have the least regional GDP inequality, seeing as they tend
to only consist of one region, and thus come out with a calculated Gini
of zero.

Regions with larger labour forces tend to show wider internal GDP
variation. This is consistent with regions that contain both highly
productive sub-areas and weaker ones. Policymakers in these regions may
need to invest specifically in connecting the smaller municipalities to
the dominant ``key region'' through commuting infrastructure, business
support, or coordinated urban planning.

The unemployment segmentation reinforces the point that ``low
disparities'' are not always a sign of strength. Regions with weaker
economic activity often show very low internal differences simply
because no municipality stands out as a strong performer. Policymakers
should avoid interpreting these cases as success stories and instead
view them as indicators of stagnation.

Finally, the panel results show that once we control for each region's
structural characteristics, the effect of growth on internal disparities
barely changes. The implication is: regional GDP inequality is mainly
driven by long-term spatial and economic structures, not by short-term
GDP growth. Effective policies therefore need to focus on durable
place-based strategies---strengthening urban centres, improving
connectivity, and reducing structural disadvantages---rather than
relying on general growth policies to equalise development inside
regions.

\subsection{Limitations and Future Research -
Harald}\label{limitations-and-future-research---harald}

\begin{itemize}
\tightlist
\item
  lack of data
\item
  What variable to use
\item
  lack of information on what variables that may have a bigger impact on
  regional inequalities.
\item
  Different countries, different ways of collecting data
\item
  Gain more insight in each country and gain a more understanding on
  collecting data.
\end{itemize}

\subsubsection{Reasearch limitations}\label{reasearch-limitations}

\subsection{Conclusion}\label{conclusion}

\subsubsection{Summary}\label{summary}

\subsubsection{Final Reflection}\label{final-reflection}

\subsection{References}\label{references}

\subsection*{Appendix}\label{appendix}
\addcontentsline{toc}{subsection}{Appendix}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-countyi2018}
\emph{County incomes and regional GDP 2015 - CSO - central statistics
office}. (2018). CSO.
\url{https://www.cso.ie/en/releasesandpublications/er/cirgdp/countyincomesandregionalgdp2015/}

\bibitem[\citeproctext]{ref-informat}
\emph{Information on data - national accounts - eurostat}. (n.d.).
\url{https://ec.europa.eu/eurostat/web/national-accounts/information-data}

\bibitem[\citeproctext]{ref-lessmann2017}
Lessmann, C., \& Seidel, A. (2017). Regional inequality, convergence,
and its determinants -- a view from outer space. \emph{European Economic
Review}, \emph{92}, 110--132.
\url{https://doi.org/10.1016/j.euroecorev.2016.11.009}

\end{CSLReferences}




\end{document}
